Locality and Effect Reflection

### 2025-07-09 paper.tex
Maybe a first draft?

\paragraph{Algebraic effects} Algebraic effects and handlers are an
increasingly popular approach to programming with effects. Handlers
allow the programmer to add additional effects to the \emph{ambient
  monad} of their language -- the monad describing which effects an
ordinary function can perform -- for some region of their program.

@@

\paragraph{Locality and effect reflection}
Locality captures the general idea of a value depending on some notion
of ``location''. We can represent it in a type system using a comonadic
\emph{global modality} $\glob$ that tracks which values do not depend on
their location. A particularly useful notion of ``location'' is the
ambient monad.

@@

\paragraph{Effect reflection in OxCaml}
OxCaml\cite{lorenzen2024oxidizing} provides type system support for
tracking locality and runtime support for algebraic effects. For
backwards compatibility with OCaml, an unadorned arrow type
\lstinline[style=oxcaml]{t -> s} is treated as if the parameter and
return types were actually \lstinline[style=oxcaml]{[]t} and
\lstinline[style=oxcaml]{[]s}. To get parameters or results that are not
implicitly global, we write
e.g. \lstinline[style=oxcaml]{t @ local -> s @ local}.

@@

\todo{Add some mention of connection to lexical and named handlers}

\paragraph{Denotational semantics} The design of effect reflection falls
out naturally from a simple denotational semantics of algebraic
effects. The underlying idea is to index the semantics by the ambient
monad. Rather than the semantics of a type being a set it is instead a
function from ambient monads to set: if you tell it what the ambient
monad is it can tell you what its set of values is.

@@

\item The semantics of our handler types $Y$ are given by the Yoneda
  embedding, and our effect reflection isomorphism becomes a simple
  consequence of the Yoneda lemma.
\end{itemize}

\paragraph{Structure of the paper} We proceed as follows:
\begin{itemize}
\item In Section \ref{sec:pracice}, we give an overview of effect reflection in
  practice in OxCaml
\item In Section \ref{sec:semantics}, we describe the constructions for a
  novel denotational semantics of algebraic effects
\item In Section \ref{sec:language}, we define a simply typed calculus featuring
  effect reflection along with its denotational semantics
\item In Section \ref{sec:implementation}, we give an implementation of effect
  reflection in OxCaml in terms of the untracked algebraic effects it
  inherits from OCaml.
\item In Section \ref{sec:related-work}, we discuss related work, and in
  Section \ref{sec:future-work}, we discuss future work.
\end{itemize}

\section{Effect reflection in practice}
\label{sec:pracice}

@@

module Handler(O : Op) : sig
  type t
end
\end{lstlisting}
For an effect signature \lstinline[style=oxcaml]{O},
\lstinline[style=oxcaml]{'a Term(O).t} is the type of terms in the free
algebra of the corresponding effect -- otherwise known as the freer
monad\cite{??}. \lstinline[style=oxcaml]{Handler(O).t} is the type of
mappings from the operations of \lstinline[style=oxcaml]{O} into the
operations of the current ambient monad. Alternatively,
\lstinline[style=oxcaml]{Handler(O).t} can be described as representing
algebraic effect handlers that are handling the current computation --
hence the name.

@@

This shows how to implement state using effect reflection, but note that
the same interface can also be implemented direcly using mutable state
and locality:
\begin{lstlisting}[style=oxcaml]
type handler = int ref

@@

This transformation from effect relection to local mutable state works
for any \emph{tail-resumptive} effect handler.

@@

To acheive this more general form of composition, we require an
additional capability in our effect reflection API. We need to be able
to manipulate \lstinline[style=oxcaml]{Handler} types for sums of
effects.
\begin{lstlisting}[style=oxcaml]
module Sum (L : Op) (R : Op) : sig
  type 'a t =
    | Left : 'a L.t -> 'a t
    | Right : 'a R.t -> 'a t
end

@@

The literature on lexical effect handlers often ignores this issue. For
example, \cite{??} proposed a system which only supported directly
nesting effect handlers and had no mechanism for writing a composable
handler for generators that could be moved between different underlying
asynchronous I/O handlers.

@@

The need for a new copy of the interface for each type constructor arity
is an artifact of OCaml's approach to higher-kinded types. Languages
with direct support for higher-kinded types should be able to support
parameterized effects of arbitrary arity with a single interface.

@@

Effect reflection requires, for any signature $\Sigma$ that is being
reflected, an inductive type $\ind{\Sigma}$ to represent the free
algebra of $\Sigma$. $\ind{\Sigma}$ is sometimes called the \emph{freer
  monad} of $\Sigma$ \cite{??}.

As with other algebraic effects systems in the literature, we only
support handlers for algebraic theories without any equations, even
though our semantics is defined in terms of arbitrary theories. We
define algebraic signatures $\Sigma$ as finite sets of operations:
\begin{equation*}
\{ op_1 \types \tau_{p_1} \rightarrowtail \tau_{a_1}; \ldots; op_n
\types \tau_{p_n} \rightarrowtail \tau_{a_n} \}
\end{equation*}
An operation type $\tau_p \rightarrowtail \tau_a$ indicates an operation
parameterised by $\tau_p$ and with arity $\tau_a$.

@@

Note that the reduction relations do not include $\eta$ expansion. They
also do not include some conversions you might expect for computations:
\begin{itemize}
\item There is no rule to convert
  $\; \letv{x_2}{\letv{x_1}{c_1}{c_2}}{c_3} \,$ to
  $\; \letv{x_1}{c_1}{\letv{x_2}{c_2}{c_3}} \,$ or vice versa.
\item There is no rule to convert $\; \letv{x}{c}{\return x} \,$ to $c$.
\end{itemize}
These rules are excluded because they are not $\beta$ rules. Instead
they are better thought of as $\nu$ rules \cite{??}: additional
equalities between neutral terms that are immediate consequences of
structural induction. Rather than be included as part of ordinary
reduction they can be done as an additional standardization pass on
neutral terms after reduction and $\eta$-expansion have completed. This
staged approach to computing equality of terms works because $\nu$ rules
cannot produce additional $\beta$ redexes.

@@

To this end, we will restrict the ambient monad to be an \emph{algebraic
  monad}: one that can be expressed as the free algebras of an algebraic
theory. Note that this defintion of algebraic monad is not precise:
different choices for the definition of ``algebraic theory'' give you
different meanings for ``algebraic monad''. We'll make our particular
choice over the course of this section.

@@

Arities in algebraic signatures are often considered as natural numbers,
but they can equivalently be just sets, with families of terms indexed
by that set used as the arguments of that operator in a term. It is also
convenient to allow families of operations parameterized by some set. As
such, we can present a signature in the form:
\begin{equation*}
\{ op_1 \types P_1 \rightarrowtail A_1; \ldots; op_n
\types P_n \rightarrowtail A_n \}
\end{equation*}
where $op \types P \rightarrowtail A$ indicates an operation $op$
parameterised by $P$ and with arity $A$.

@@

Most monads used in practice for programming are algebraic by this
definition, with the notable exception of the continuation monad. The
issue with trying to represent non-algebraic monads algebraically is
essentially one of size: if you try to write down the collection of
algebirac operations for the continuation monad you get something that
is too large to be a set. This is closely related to the strict
positivity restriction on inductive type definitions.

@@

\begin{definition}
  A \emph{Lawvere theory} is a category $L$ with finite products in
  which every object is isomorphic to some finite cartesion product
  $x^n = x \times \ldots \times x$ of a distinguished object $x$.
\end{definition}
The objects of a Lawvere theory correspond to the different possible
arities and the morphisms $x^n \rightarrow x$ correspond to terms
of $L$ with $n$ free variables quotiented by the equations in $L$. A
model of a Lawvere theory $L$ is a product-preserving functor
$L \rightarrow \setc$.

@@

Within our semantics, this amounts to restricting a type to only the
elements which do not depend on the ambient monad. This is very similar
to the purity comonad in ??\cite{??}. In particular,
\begin{equation*}
  \glob \mon(P) \quad = \quad T_{\initial} \circ \glob P
  \quad \cong \quad \glob P
\end{equation*}
as elements of the free algebras of the initial algebraic theory are
just elements of their carrier sets.

@@

The essence of algebraic effect handlers is taking a computation in
$T_{L + \Sigma}(S)$ the free algebras of the sum of an algebraic theory
$L$ and an algebraic signature $\Sigma$, spliting out and handling
$\Sigma$, leaving a computation in only the algebraic theory
$T_{L}(S^\prime)$.

The key to this process is the following lemma
\cite{hyland2006combining}.
\begin{lemma}
\label{lemma:sums}
  For any Lawvere theory $L$ and algebraic signature $\Sigma$
  \begin{equation*}
  T_{L + \Sigma}(S) \cong \mu X.T_L(S + F_\Sigma(X))
  \end{equation*}
\end{lemma}
where $\mu X.G(X)$ is the intial algebra of some polynomial endofunctor
$G$, and $F_\Sigma$ is the endofunctor on $\setc$ for which $T_{\Sigma}$
is the free monad.

Effect reification works by exposing $\mu X.T_L(S + F_\Sigma(X))$ to the
user via an inductive type. Categories of presheaves have initial
algebras of polynomial endofunctors, i.e. W-types, so we can construct
the required initial algebra as a presheave. Given an algebraic
signature $\Sigma$ and a presheave $P$, we define the following family
of initial algebras:
\begin{align*}
&\ind{\Sigma} \types [\pshlawcop, \pshlawcop] \\
&\ind{\Sigma}(P) = \mu X. (P + F_\Sigma \circ \mon{X})
\end{align*}
and observe the following variation on Lemma \ref{lemma:sums}.
\begin{lemma}
\label{lemma:sums2}
  For any Lawvere theory $L$ and algebraic signature $\Sigma$
  \begin{equation*}
  T_{L + \Sigma}(P(L)) \cong \mon{(\ind{\Sigma}(P))}(L)
  \end{equation*}
\end{lemma}

@@

Effect reflection is also closely related to \emph{named effect
  handlers}\cite{xie2022first}. Named effect handlers use higher-rank
polymorphism to emulate locality tracking in much the same way as the
\lstinline[style=haskell]{St} monad in Haskell. This still requires
using an effect system to track which named effects are used by a
computation, in order to ensure that those effects do not escape the
scope of their handler. Using higher-rank polymorphism also prevents
global type inference, requiring complex inference that extends beyond
classic Hindly-Milner, and demanding a higher annotation burden on the
programmer.

>>>>

\subsection{Algebraic effects}
Algebraic effects and handlers are an increasingly popular approach to
programming with effects. Handlers allow the programmer to add
additional effects to the \emph{ambient monad} of their language -- the
monad describing which effects an ordinary function can perform -- for
some region of their program.

@@

\subsection{Locality and effect reflection}
Locality captures the general idea of a value depending on some notion
of ``location''. We can represent it in a type system using a comonadic
\emph{global modality} $\glob$ that tracks which values do not depend on
their location. A particularly useful notion of ``location'' is the
ambient monad.

@@

\subsection{Effect reflection in OxCaml}
OxCaml\cite{lorenzen2024oxidizing} provides type system support for
tracking locality and runtime support for algebraic effects. For
backwards compatibility with OCaml, an unadorned arrow type
\lstinline[style=oxcaml]{t -> s} is treated as if the parameter and
return types were actually \lstinline[style=oxcaml]{[]t} and
\lstinline[style=oxcaml]{[]s}. To get parameters or results that are not
implicitly global, we write
e.g. \lstinline[style=oxcaml]{t @ local -> s @ local}.

@@

\subsection{Denotational semantics} The design of effect reflection falls
out naturally from a simple denotational semantics of algebraic
effects. The underlying idea is to index the semantics by the ambient
monad. Rather than the semantics of a type being a set it is instead a
function from ambient monads to set: if you tell it what the ambient
monad is it can tell you what its set of values is.

@@

\item The semantics of our handler types $\yoneda{\Sigma}$ are given by
  the Yoneda embedding, and our effect reflection isomorphism becomes a
  simple consequence of the Yoneda lemma and some properties of the sums
  of algebraic theories.
\end{itemize}

\subsection{Structure of the paper} We proceed as follows:
\begin{itemize}
\item In Section \ref{sec:practice}, we give an overview of effect reflection in
  practice in OxCaml.
\item In Section \ref{sec:language}, we define a simply typed calculus featuring
  effect reflection.
\item In Section \ref{sec:semantics}, we describe our denotational
  semantics of effect reflection.
\item In Section \ref{sec:implementation}, we give an implementation of effect
  reflection in OxCaml in terms of the untracked algebraic effects it
  inherits from OCaml.
\item In Section \ref{sec:related-work}, we discuss related work.
\end{itemize}

\section{Effect reflection in practice}
\label{sec:practice}

@@

module Handler(O : Op) : sig
  type t
end
\end{lstlisting}
For an effect signature \lstinline[style=oxcaml]{O},
\lstinline[style=oxcaml]{'a Term(O).t} is the type of terms in the free
algebra of the corresponding effect -- otherwise known as the freer
monad\cite{kiselyov2015freer}. \lstinline[style=oxcaml]{Handler(O).t} is
the type of mappings from the operations of \lstinline[style=oxcaml]{O}
into the operations of the current ambient monad. Alternatively,
\lstinline[style=oxcaml]{Handler(O).t} can be described as representing
algebraic effect handlers that are handling the current computation --
hence the name.

@@

This shows how to implement state using effect reflection, but note that
the same interface can also be implemented directly using mutable state
and locality:
\begin{lstlisting}[style=oxcaml]
type handler = int ref

@@

This transformation from effect reflection to local mutable state works
for any \emph{tail-resumptive} effect handler.

@@

To achieve this more general form of composition, we require an
additional capability in our effect reflection API. We need to be able
to manipulate \lstinline[style=oxcaml]{Handler} types for sums of
effects.
\begin{lstlisting}[style=oxcaml]
module Sum (L : Op) (R : Op) : sig
  type 'a t =
    | Left : 'a L.t -> 'a t
    | Right : 'a R.t -> 'a t
end

@@

The literature on lexical effect handlers often ignores this issue. For
example, Xie et al. \cite{xie2020effect} proposed a system which only
supported directly nesting effect handlers and had no mechanism for
writing a composable handler for generators that could be moved between
different underlying asynchronous I/O handlers.

@@

The need for a new copy of the interface for each type constructor arity
is an artefact of OCaml's approach to higher-kinded types. Languages
with direct support for higher-kinded types should be able to support
parameterized effects of arbitrary arity with a single interface.

@@

Effect reflection requires, for any signature $\Sigma$ that is being
reflected, an inductive type $\ind{\Sigma}$ to represent the free
algebra of $\Sigma$. $\ind{\Sigma}$ is sometimes called the \emph{freer
  monad} of $\Sigma$ \cite{kiselyov2015freer}.

As with other algebraic effects systems in the literature, we only
support handlers for algebraic theories without any equations, even
though our semantics is defined in terms of arbitrary theories. We
define algebraic signatures $\Sigma$ as finite sets of operations:
\begin{equation*}
\{ op_1 \types \tau_{p_1} \rightarrowtail \tau_{a_1}; \ldots; op_n
\types \tau_{p_n} \rightarrowtail \tau_{a_n} \}
\end{equation*}
An operation type $\tau_p \rightarrowtail \tau_a$ indicates an operation
parameterized by $\tau_p$ and with arity $\tau_a$.

@@

Note that the reduction relations do not include $\eta$ expansion. They
also do not include some conversions you might expect for computations:
\begin{itemize}
\item There is no rule to convert
  $\; \letv{x_2}{\letv{x_1}{c_1}{c_2}}{c_3} \,$ to
  $\; \letv{x_1}{c_1}{\letv{x_2}{c_2}{c_3}} \,$ or vice versa.
\item There is no rule to convert $\; \letv{x}{c}{\return x} \,$ to $c$.
\end{itemize}
These rules are excluded because they are not $\beta$ rules. Instead
they are better thought of as $\nu$ rules \cite{allais2013new}:
additional equalities between neutral terms that are immediate
consequences of structural induction. Rather than be included as part of
ordinary reduction they can be done as an additional standardization
pass on neutral terms after reduction and $\eta$-expansion have
completed. This staged approach to computing equality of terms works
because $\nu$ rules cannot produce additional $\beta$ redexes.

@@

To this end, we will restrict the ambient monad to be an \emph{algebraic
  monad}: one that can be expressed as the free algebras of an algebraic
theory. Note that this definition of algebraic monad is not precise:
different choices for the definition of ``algebraic theory'' give you
different meanings for ``algebraic monad''. We'll make our particular
choice over the course of this section.

@@

Arities in algebraic signatures are often considered as natural numbers,
but they can equivalently be just sets, with families of terms indexed
by that set used as the arguments of that operator in a term. It is also
convenient to allow families of operations parameterized by some set. As
such, we can present a signature in the form:
\begin{equation*}
\{ op_1 \types P_1 \rightarrowtail A_1; \ldots; op_n
\types P_n \rightarrowtail A_n \}
\end{equation*}
where $op \types P \rightarrowtail A$ indicates an operation $op$
parameterized by $P$ and with arity $A$.

@@

Most monads used in practice for programming are algebraic by this
definition, with the notable exception of the continuation monad. The
issue with trying to represent non-algebraic monads algebraically is
essentially one of size: if you try to write down the collection of
algebraic operations for the continuation monad you get something that
is too large to be a set. This is closely related to the strict
positivity restriction on inductive type definitions.

@@

\begin{definition}
  A \emph{Lawvere theory} is a category $L$ with finite products in
  which every object is isomorphic to some finite cartesian product
  $x^n = x \times \ldots \times x$ of a distinguished object $x$.
\end{definition}
The objects of a Lawvere theory correspond to the different possible
arities and the morphisms $x^n \rightarrow x$ correspond to terms
of $L$ with $n$ free variables quotiented by the equations in $L$. A
model of a Lawvere theory $L$ is a product-preserving functor
$L \rightarrow \setc$.

@@

Within our semantics, this amounts to restricting a type to only the
elements which do not depend on the ambient monad. This is very similar
to the purity comonad of Choudhury and
Krishnaswami\cite{choudhury2020recovering}. In particular,
\begin{equation*}
  \glob \mon(P) \quad = \quad T_{\initial} \circ \glob P
  \quad \cong \quad \glob P
\end{equation*}
as elements of the free algebras of the initial algebraic theory are
just elements of their carrier sets.

@@

The essence of algebraic effect handlers is taking a computation in
$T_{L + \Sigma}(S)$ the free algebras of the sum of an algebraic theory
$L$ and an algebraic signature $\Sigma$, splitting out and handling
$\Sigma$, leaving a computation in only the algebraic theory
$T_{L}(S^\prime)$.

The key to this process is the following lemma
\cite{hyland2006combining}.
\begin{lemma}
\label{lemma:sums}
  For any Lawvere theory $L$ and algebraic signature $\Sigma$
  \begin{equation*}
  T_{L + \Sigma}(S) \cong \mu X.T_L(S + F_\Sigma(X))
  \end{equation*}
\end{lemma}
where $\mu X.G(X)$ is the initial algebra of some polynomial endofunctor
$G$, and $F_\Sigma$ is the endofunctor on $\setc$ for which $T_{\Sigma}$
is the free monad.

Effect reification works by exposing $\mu X.T_L(S + F_\Sigma(X))$ to the
user via an inductive type. Categories of presheaves have initial
algebras of polynomial endofunctors, i.e. W-types, so we can construct
the required initial algebra as a presheaf. Given an algebraic
signature $\Sigma$ and a presheaf $P$, we define the following family
of initial algebras:
\begin{align*}
&\ind{\Sigma} \types [\pshlawcop, \pshlawcop] \\
&\ind{\Sigma}(P) = \mu X. (P + F_\Sigma \circ \mon{X})
\end{align*}
and observe the following variation on Lemma \ref{lemma:sums}.
\begin{lemma}
\label{lemma:sums2}
  For any Lawvere theory $L$ and algebraic signature $\Sigma$
  \begin{equation*}
  T_{L + \Sigma}(P(L)) \cong \mon{(\ind{\Sigma}(P))}(L)
  \end{equation*}
\end{lemma}

@@

Effect reflection is also closely related to \emph{named effect
  handlers}\cite{xie2022first}. Named effect handlers use higher-rank
polymorphism to emulate locality tracking in much the same way as the
\lstinline[style=haskell]{St} monad in Haskell. This still requires
using an effect system to track which named effects are used by a
computation, in order to ensure that those effects do not escape the
scope of their handler. Using higher-rank polymorphism also prevents
global type inference, requiring complex inference that extends beyond
classic Hindley-Milner, and demanding a higher annotation burden on the
programmer.
